Characteristic polynomials of p-adic matrices

### 2017-01-30 charpoly/charpoly.tex
Working on section 2

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\R}{\mathbb R}
\newcommand{\OK}{\mathcal{O}_K}

@@

\subsection{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer.

@@

The main benefit of working with lattices is the following result
\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.  For a matrix $M \in M_n(K)$
we identify the tangent space $V$ at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.  Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed)
ball of radius $r$ in $V$.  
Our analysis of the $\chi$ begins with the following result
\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ and that $d\chi(M): V \to W$ is surjective.
Let $C$ be a positive real number such that
$B_W(1) \subset d\chi(M)(B_V(C))$.  Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

\subsection{Computation of the differential}

In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi(M)$ is given
by
\[
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\]
Moreover, $d\chi(M)$ will be surjective exactly when the Jordan blocks of $M$ all have distinct generalized eigenvalues.

\todo{Formula with comatrix.}

@@

\todo{Surjectivity. Stability under multiplication by $X$ mod $\chi$.}

@@

\section{Optimal precision\\via Hessenberg form}
%\section{Computation of the characteristic polynomial and its derivative}

@@

\begin{prop}
Let $cond(M)$ be the valuation of the last invariant factor of $M.$
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal
at least when $N$ is large.
\end{prop}
\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
It is $-2cond(M)$ for $M^{-2}$ and it is then clear that $-2cond(M)$
can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
and the smallest that can be achieved this way for $dM$ in a precision lattice
of flat precision. Hence the optimality of the bound given, at least when 
$N$ is large, thanks to Lemma 3.14 of \cite{caruso-roe-vaccon:14a}

@@

@@

When this time $dM$ is jagged, the trick of the proof of Proposition
\ref{prop:optimal_flat} can not work anymore.
Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
are not easy, as $P^{-1}  dM P$ is not jagged anymore.
Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
conclude that optimal precision can still be read in $\softO (n^3)$ 
operations in $O_K.$

%Thanks to the previous results, we can now compute 
%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
%using the following algorithm.

@@

\section{Optimal precision\\via Frobenius form}

\section{Eigenvalues}

\subsection{Optimal precision on a simple eigenvalue}

\todo{Optimal precision on $P(\lambda_i)$ and on $\lambda_i$}

@@

%\subsection{Algorithm}

>>>>

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\R}{\mathbb R}

@@

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}
Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

\begin{proof}
We first show that the distinct generalized eigenvalues condition
guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
matrices with a single $1$ entry and all other entries $0$, it suffices to show
that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
are those on or below the diagonal within the Jordan blocks, and within the $i$th
block, each entry along the $j$th subdiagonal is
$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
As long as the $\lambda_i$ are distinct, these form a basis for $W$.

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
Since the precision polygon lies below the Newton polygon \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
and since the smallest entry of the Newton polygon occurs either with the leading
coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(C))$.

Since the coefficients of $\chi(M)$ are given by polynomials in the entries of $M$
with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
the conclusion.
\end{proof}

@@

Consider the isomorphism $W \to K[X] / \chi(X)$ induced by inclusion.
\begin{proposition}
Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
\end{proposition}
\begin{proof}
\todo{Finish proof}
\end{proof}

We can also give a criterion on $M$ that implies 
\begin{proposition}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
\end{proposition}
\begin{proof}
First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
\end{proof}

@@

\section{Computation of the characteristic polynomial and its derivative}

@@

\begin{prop}
Let $cond(M)$ be the valuation of the last invariant factor of $M.$
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal
at least when $N$ is large.
\end{prop}
\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
It is $-2cond(M)$ for $M^{-2}$ and it is then clear that $-2cond(M)$
can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
and the smallest that can be achieved this way for $dM$ in a precision lattice
of flat precision. Hence the optimality of the bound given, at least when 
$N$ is large, thanks to Lemma 3.14 of \cite{CRV:2014}

@@

%\subsection{$\com (X -M)$ and jagged precision}

\subsection{Factorization}

When this time $dM$ is jagged, the trick of the proof of Proposition
\ref{prop:optimal_flat} can not work anymore.
Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
are not easy, as $P^{-1}  dM P$ is not jagged anymore.
Fortunately, a factorization result of $\com (X-M)$ will be enough
to conclude that optimal precision can still be read in $\softO (n^3)$
operations in $O_K.$

@@

Thanks to the previous results, we can now compute 
$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
using the following algorithm.

@@

\section{The case of diagonalizable matrices}

@@

\subsection{Algorithm}
